---
title: "Pstat 131 Final Project: Predicting IPO First-Day \"Pops\""
author: "Andy"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: cosmo
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.align = "center")
```

# Introduction

## Background

For my project I choose to pick whether an initial public offering (IPO) will "pop".
An IPO is the first time a private company sells
shares of stock to the public on an open exchange like the NASDAQ. 
Before the trading begins, the company and the investment banks that manage the
offering agree on an offer price which is the price at which shares are
offered to investors before the stock starts trading. Once the offering
opens the stock up to the public, the price often moves quickly away
from that offer price. When a stock's price jumps above its offer
price on its first day of trading, it has "popped" and
financial investors refer to this as IPO underpricing:
the company effectively sold its shares for less than the market was
immediately willing to pay for them. When the stock instead trades the same price or
falls below its offer price it means the IPO was valued properly or overvalued.

Predicting whether an IPO will pop is a common question in
financial economics. Previous research has linked underpricing to factors like
the company's industry sector, the general condition of the stock market
leading up to the offering, and things like company age and
size. For this project I will build a supervised classification model that
uses only information that would have been known before or at the moment
a company's shares begin trading like sector, company age, employee count,
whether the company is U.S.-based, and how the market had been
trending in the months before the IPO to predict a binary outcome of whether
 the stock "pop" on its first day of trading or not?

## Why this project

I choose this project because it recently aligns with my internship of financial
data analyst were I was analyzing market data for an insurance company that insures
broker-dealers. Broker-Dealers are companies that help investors buy and sell stocks
basically like a middleman for example Robinhood or Webull. Seeing whether IPO's
would "pop" or not was interesting to me because usually when an IPO is offered it is
offered trough an Investment Banker like Morgan Stanley but with the new stock market
you could buy it through these middleman like Robinhood. I found this to be a useful
project idea since it is such a new topic and finding out if IPO's would "pop"
would be beneficial information for these insurance companies since there is more 
risk now that these middlemen can sell IPO's. 

## Project roadmap

The rest of this report follows the basic ML pipeline. First I
load and clean the raw IPO data and make the binary response variable
`pop`. Then I look at the data visually to understand the outcome and
relationship to the candidate predictors. I then split the data into
training and testing sets and set up stratified k-fold cross-validation on
the training set. Nexy fitting and tuneing the  five different types of classification
models using the same recipe and the same folds. I will then choose the
best-performing model based on cross-validated ROC AUC, fit it to the full
training set, and look at its performance one time on the held-out testing
set.

# Data

## Source and citation

**Source:** "Stocks IPO information & results," Kaggle dataset by user
`proselotis`. Available at:
<https://www.kaggle.com/datasets/proselotis/financial-ipo-data>

**Citation:** Proselotis. (2020). *Stocks IPO information & results*
[Data set]. Kaggle. https://www.kaggle.com/datasets/proselotis/financial-ipo-data

The raw file (`IPODataFull.csv`) contains 3,762 U.S. IPOs listed between
1996 and 2018 which has one row per company. For each IPO it includes the company's
sector/industry classification, basic firm characteristics (employee count,
founding year, CEO/president information), the broader market's trend in
the months leading up to the listing, and daily open/high/low/close/volume
trading prices for around 260 trading days following the IPO also alongside
several outcome measures the original author had already made. The raw file is 
extremely wide because it includes a full day-by-day price history and a
large block of line-item financial-statement data for the fiscal year
before each company's IPO so I will only keep the small subset of columns relevant
to my project.

## Loading the raw data

```{r load-packages}
library(tidyverse)
library(tidymodels)
library(janitor)
library(corrplot)
library(discrim)   
library(vip)        
library(kknn)
library(themis)  
library(doParallel)
tidymodels_prefer()
```

```{r load-data}
ipo_raw <- read_csv("data/IPODataFull.csv", locale = locale(encoding = "latin1")) %>%
  clean_names()   # snake_case everything, same as in Homework 5

dim(ipo_raw)
```

The `clean_names()` function from `janitor` turns every column name into
`snake_case` so I don't have to remember the exact capitalization later.
The file is saved in a Latin-1 encoding instead of UTF-8 because some of
the company names have special characters in them, so I have to tell
`read_csv()` that or it throws parsing errors. Knitting this chunk also
gives a "one or more parsing issues" warning. I checked it with
`problems()` and all of the flagged cells are in the giant block of
per-industry, pre-IPO financial columns (like a column that's supposed to
be a boolean flag but sometimes has an actual dollar amount in it
instead). I don't use any of those columns in the next step, so this
warning is safe to ignore for this project.

## Selecting and cleaning the variables I need

Out of the ~1,660 raw columns, two of them give me exactly the response
variable I need `open_day0` and `close_day0`. When I looked closely at
the raw file I noticed `open_day0` actually acts like the **IPO offer price** 
instead of the first price that actually got traded. For a lot of
rows the day's low is actually above `open_day0`, which can only happen
if the stock started trading above its offer price right away and never
came back down to it. `close_day0` is just the stock's closing price at
the end of its first day of trading. Offer price compared to first-day
close is actually a pretty standard way IPO underpricing gets measured in
finance research whcih works out well for this project.

For the predictors, I picked variables that describe the company and the
market conditions of the IPO, and I left out a few of the raw
columns `market_cap`, `last_sale`, `revenue`, `net_income`, and the raw
`employees` count because they turned out to be snapshots of the company
from whenever the dataset was taken in 2020 instead of from the
actual IPO date. Using those would leak information about the company's
future into the model.

```{r select-clean}
ipo <- ipo_raw %>%
  transmute(
    symbol             = symbol,
    ipo_year           = year,
    sector             = sector,
    usa_company        = usa_company,
    employees_grouped  = employees_grouped,
    year_difference    = year_difference,     # company age (yrs) at IPO
    ceo_gender         = usable_ceo_gender,
    market_month_trend  = market_month_trend,   # market return, 1 month before IPO
    market_3month_trend = market3month_trend,   # market return, 3 months before IPO
    market_6month_trend = market6month_trend,   # market return, 6 months before IPO
    market_year_trend   = market_year_trend,    # market return, 1 year before IPO
    offer_price        = open_day0,
    close_day0         = close_day0
  ) %>%
  mutate(
    employees_grouped = employees_grouped %>% replace_na("Unknown") %>%
      na_if("Blank") %>% replace_na("Unknown"),
    ceo_gender = ceo_gender %>% replace_na("Unknown") %>%
      recode("Blank" = "Unknown", "unknown" = "Unknown", "andy" = "Unknown"),
    usa_company = usa_company %>% replace_na("Unknown"),
    pct_change = (close_day0 - offer_price) / offer_price
  ) %>%
  filter(!is.na(offer_price), !is.na(close_day0), offer_price > 0) %>%
  mutate(
    sector = str_replace_all(sector, "[^A-Za-z0-9]+", "_"),
    employees_grouped = str_replace_all(employees_grouped, "[^A-Za-z0-9]+", "_")
  ) %>%
  mutate(across(c(sector, usa_company, employees_grouped, ceo_gender), as.factor))

glimpse(ipo)
```

# Exploratory Data Analysis

## Missing data

```{r missing-data, fig.width=6, fig.height=4}
missing_summary <- ipo %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") %>%
  mutate(pct_missing = n_missing / nrow(ipo)) %>%
  filter(n_missing > 0) %>%
  arrange(desc(n_missing))

missing_summary

ggplot(missing_summary, aes(x = fct_reorder(variable, pct_missing), y = pct_missing)) +
  geom_col(fill = "#e34a33") +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Proportion of Missing Values by Column",
       x = NULL, y = "% missing") +
  theme_minimal()
```

`sector` and `year_difference` which are the company age at IPO, are the two predictors
with real amounts of missing data whcih is around 19% and 20%. This lines up with
what I said in my data memo could be a risk for smaller or older listings
where the company's basic info was never fully recorded. I decided not to
drop either column since 1/5 missing isn't really that bad.
Instead in the recipe below I add an explicit
`"unknown"` category for missing `sector` values using `step_unknown()`,
and use the median for the missing `year_difference` values using
`step_impute_median()`. I also noticed there were 63 duplicated ticker
symbols in the raw data, which is probably just a ticker getting reused
years apart by two different companies. Since each row is still its own
distinct IPO event with its own date and prices I decided to just leave
these rows in instead of removing them.

## Defining "pop"

```{r define-pop, fig.width=6, fig.height=4}
ggplot(ipo, aes(x = pct_change)) +
  geom_histogram(bins = 60, fill = "#2c7fb8", color = "white") +
  geom_vline(xintercept = 0.10, linetype = "dashed", color = "firebrick") +
  coord_cartesian(xlim = c(-0.25, 1)) +
  scale_x_continuous(labels = scales::percent) +
  labs(title = "Distribution of First-Day IPO Returns",
       subtitle = "Dashed line marks the 10% 'pop' threshold used below (x-axis trimmed for readability)",
       x = "First-day return: (close, day 0 - offer price) / offer price",
       y = "Number of IPOs") +
  theme_minimal()

quantile(ipo$pct_change, probs = c(0.5, 0.75, 0.9, 0.95, 0.99), na.rm = TRUE)
```

The distribution of first-day returns is really right-skewed. The median
IPO in this dataset only gains about 2% on its first day wuth the top 10%
gaining 14% or more and a handful of low-priced IPOs gaining several
hundred percent. I decided to define an IPO as having "popped" if its
first-day return is over 10%, which is a threshold that roughly
separates the top ~15% of IPOs from the rest. This is also consistent with 
the underpricing thresholds I saw used in typical financial data I read for my data memo.

```{r create-outcome}
ipo <- ipo %>%
  mutate(pop = if_else(pct_change > 0.10, "Pop", "No_Pop")) %>%
  mutate(pop = factor(pop, levels = c("Pop", "No_Pop")))

ipo %>% count(pop) %>% mutate(prop = n / sum(n))
```

As I expected a stock "popping" is the minority class at only around 15%
of IPOs. This confirms the class-imbalance risk I flagged in my data memo,
which is why I stratify every split and every resampling fold on `pop`
below. This is also why I mainly used ROC AUC instead of accuracy since ROC AUC
is a lot less sensitive to class imbalance.

## Pop rate by sector

```{r eda-sector, fig.width=7, fig.height=5}
ipo %>%
  mutate(sector = fct_na_value_to_level(sector, "Unknown")) %>%
  group_by(sector) %>%
  filter(n() >= 30) %>%   # only show sectors with a reasonable sample size
  summarise(pop_rate = mean(pop == "Pop"), n = n()) %>%
  ggplot(aes(x = fct_reorder(sector, pop_rate), y = pop_rate)) +
  geom_col(fill = "#2c7fb8") +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "First-Day Pop Rate by Sector",
       subtitle = "Sectors with fewer than 30 IPOs in the data excluded",
       x = NULL, y = "% of IPOs that popped") +
  theme_minimal()
```

There's real spread between sectors here. Technology and Miscellaneous IPOs
pop about 28% of the time when compared to only about 8% for Finance and 10%
for Public Utilities. This lines up with the "hot sector" idea from the
finance data and makes `sector` an actually useful signal for the model.

## Company age and employee count vs. outcome

```{r eda-age-boxplot, fig.width=6, fig.height=4}
ggplot(ipo, aes(x = pop, y = year_difference, fill = pop)) +
  geom_boxplot() +
  coord_cartesian(ylim = c(0, 50)) +
  scale_fill_manual(values = c("Pop" = "#31a354", "No_Pop" = "#969696")) +
  labs(title = "Company Age at IPO by Outcome",
       x = NULL, y = "Years between founding and IPO") +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r eda-employees-bar, fig.width=7, fig.height=4}
ipo %>%
  mutate(employees_grouped = fct_relevel(employees_grouped,
           "0_100", "101_500", "501_1000", "1001_5000",
           "5001_10000", "_10000", "Unknown")) %>%
  mutate(employees_grouped = fct_recode(employees_grouped,
           "0-100" = "0_100", "101-500" = "101_500", "501-1,000" = "501_1000",
           "1,001-5,000" = "1001_5000", "5,001-10,000" = "5001_10000",
           ">10,000" = "_10000")) %>%
  group_by(employees_grouped) %>%
  summarise(pop_rate = mean(pop == "Pop"), n = n()) %>%
  ggplot(aes(x = employees_grouped, y = pop_rate)) +
  geom_col(fill = "#2c7fb8") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "First-Day Pop Rate by Employee-Count Bucket",
       x = "Employees at IPO", y = "% of IPOs that popped") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

The relationship with company age ends up being pretty mixed instead of
clean. The median company that popped was actually a little bit older
at IPO of being 7 years old than one that didn't pop around 5 years old,
even though the mean age is actually lower for the ones that popped
once you average in a handful of really old companies. 
Employee count is similarly messy since mid-sized companies
which had 5,001-10,000 employees pop the most unlike the smallest ones that i
though would. Neither of these variables shows a clean pattern I was
expecting which I thought was actually useful finding on
its own. It suggests firm size and age by themselves are pretty weak
predictors, and whatever signal they do carry probably comes from how the
model combines them with sector and market conditions instead of from
either variable alone.

## Pop rate by IPO year and market trend

```{r eda-year, fig.width=7, fig.height=4}
ipo %>%
  group_by(ipo_year) %>%
  summarise(pop_rate = mean(pop == "Pop"), n = n()) %>%
  ggplot(aes(x = ipo_year, y = pop_rate)) +
  geom_col(fill = "#2c7fb8") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "First-Day Pop Rate by IPO Year",
       x = NULL, y = "% of IPOs that popped") +
  theme_minimal()
```

There's a big spike right at the peak of the dot-com run-up, with about
32% of 1999 IPOs and 33% of 2000 IPOs popping compared to roughly 15% in a
normal year. This is then followed by a sharp drop once the bubble burst under 9% of
both 1998 and 2002 IPOs popped. Outside of that specific period the
year-to-year pattern gets a lot noisier so I included the
continuous `market_*_trend` variables as predictors instead of just
relying on a rough per-year average. They let the model pick up on market
conditions in more detail than just "which year did this IPO happen in."

## Correlation among numeric predictors

```{r eda-corr, fig.width=6, fig.height=6}
ipo_num <- ipo %>%
  select(year_difference, market_month_trend, market_3month_trend,
         market_6month_trend, market_year_trend) %>%
  drop_na()

corrplot(cor(ipo_num), method = "circle", type = "upper",
          tl.col = "black", tl.srt = 45, addCoef.col = "black")
```

The four market-trend variables are fairly strongly correlated with each other. 
Correlations range from about 0.48 (1-month
vs. 1-year trend) all the way up to 0.90 (6-month vs. 1-year trend), which
makes sense because a market that was trending up over the past month 
usually also trends up over the past year. `year_difference` on the
other hand is basically uncorrelated with all four of them with every
correlation under 0.02. None of these correlations are strong enough that
I'm worried about multicollinearity messing up the linear models below.

# Data Splitting and Cross-Validation

Before I fit any models I split the cleaned data into a **training set**
and a **testing set**. The training set gets used for everything up
through model selection. I use an 80/20 split that is **stratified** 
on the outcome variable `pop`. Since "popping" is a fairly rare outcome at about 
15% of IPOs, a normal random split could easily end up with pretty different 
pop rates in the training and testing sets just by chance. Stratifying makes 
sure both sets end up with about the same proportion of pops as the whole dataset.

```{r split}
set.seed(131)
ipo_split <- initial_split(ipo, prop = 0.80, strata = pop)
ipo_train <- training(ipo_split)
ipo_test  <- testing(ipo_split)

dim(ipo_train)
dim(ipo_test)
```

Inside the training set, I use **10-fold cross-validation** to tune and
compare models before I ever touch the testing set. With k-fold
cross-validation the training data gets randomly split into k=10
roughly equal-sized groups called "folds." The model gets fit 10
separate times, and each time one fold gets held out as a mini validation
set while the model trains on the other k − 1 folds, then gets scored on
the fold that was held out. Averaging the performance across all k rounds
gives a more reliable idea of how well a model generalizes compared to just doing one
train/validation split. Just like with the initial split, I stratify the folds on 
`pop` so every fold has a fair share of pops in it.

```{r cv-folds}
ipo_folds <- vfold_cv(ipo_train, v = 10, strata = pop)
```

# Recipe

All five models below use the same exact recipe so that any differences in
performance between them come from the models themselves.

```{r recipe}
ipo_recipe <- recipe(
  pop ~ sector + usa_company + employees_grouped + year_difference +
    ceo_gender + market_month_trend + market_3month_trend +
    market_6month_trend + market_year_trend,
  data = ipo_train
) %>%
  step_unknown(sector) %>%                    
  step_impute_median(year_difference) %>%
  step_novel(sector, usa_company, employees_grouped, ceo_gender) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%                
  step_normalize(all_numeric_predictors()) %>%
  step_upsample(pop, over_ratio = 1)           

ipo_baked <- prep(ipo_recipe) %>% bake(new_data = NULL)
dim(ipo_baked)
head(ipo_baked)
```

`step_unknown()` gives the missing `sector` values their own category
instead of quietly guessing a value for them. `step_impute_median()` fills
in the roughly 20% of missing `year_difference` values with the training
set's median company age. `step_novel()` protects against a category
showing up in a testing fold that was never seen in training.
`step_dummy()` one-hot encodes the four categorical predictors.
`step_zv()` gets rid of any dummy columns that end up constant. And
`step_normalize()` centers and scales the numeric predictors, which
especially matters for the elastic net model.

**On `step_upsample()`:** I actually tried fitting a random forest both
with and without this step to see if it was worth including, instead of
just adding it in because my data memo mentioned it as something I might
need. Without upsampling, the model's cross-validated ROC AUC was a little
higher but on the testing set it predicted "No_Pop" for every single
IPO and correctly caught 0 of the 110 real pops in the testing set. This
happened because so few training examples were labeled "Pop" that the
model never actually learned to push any predicted probability above the
default 0.5 cutoff. With `step_upsample()` duplicating the "Pop" rows in the
training data until the two classes are balanced the ROC AUC barely changed
but the model actually started catching a decent chunk of real pops
around a fourth of them. `step_upsample()`only used the training data 
inside each resample so the testing set and each validation fold stay at their 
real proportions. This means it isn't inflating the performance numbers instead
it's just giving the model an actual chance to learn what the minority class looks like.

# Model Fitting

I fit and tuned five different classification model types on the same
folds using logistic regression and LDA as the simple baselines. Then used 
elastic net logistic regression, random forest, and boosted tree as the more
flexible options. I used ROC AUC as my main comparison metric
since it looks at how the model ranks predicted probabilities
across every possible classification threshold and it's a lot less
sensitive to class imbalance than plain accuracy is. That matters a lot
here since only about 15% of IPOs actually pop.

```{r check-cache}
models_file <- "data/ipo_tuned_models.rda"
models_cached <- file.exists(models_file)
if (models_cached) load(models_file)
```
## Baseline: Logistic Regression

```{r log-reg, eval=!models_cached}
log_mod <- logistic_reg() %>%
  set_engine("glm") %>%
  set_mode("classification")

log_wflow <- workflow() %>%
  add_model(log_mod) %>%
  add_recipe(ipo_recipe)

log_res <- fit_resamples(
  log_wflow,
  resamples = ipo_folds,
  metrics = metric_set(roc_auc, accuracy),
  control = control_resamples(save_pred = TRUE)
)
```

## Baseline: Linear Discriminant Analysis

```{r lda, eval=!models_cached}
lda_mod <- discrim_linear() %>%
  set_engine("MASS") %>%
  set_mode("classification")

lda_wflow <- workflow() %>%
  add_model(lda_mod) %>%
  add_recipe(ipo_recipe)

lda_res <- fit_resamples(
  lda_wflow,
  resamples = ipo_folds,
  metrics = metric_set(roc_auc, accuracy),
  control = control_resamples(save_pred = TRUE)
)
```

## Elastic Net Logistic Regression

```{r elastic-net-setup}
en_mod <- logistic_reg(penalty = tune(), mixture = tune()) %>%
  set_engine("glmnet") %>%
  set_mode("classification")

en_wflow <- workflow() %>%
  add_model(en_mod) %>%
  add_recipe(ipo_recipe)

en_grid <- grid_regular(
  penalty(range = c(-4, 0)),   # log10 scale, so 0.0001 to 1
  mixture(range = c(0, 1)),
  levels = 10
)
```

```{r compute faster}
set.seed(131)
cl <- makePSOCKcluster(parallel::detectCores() - 1)
registerDoParallel(cl)
```


```{r elastic-net-tune, eval=!models_cached}
en_tune <- tune_grid(
  en_wflow,
  resamples = ipo_folds,
  grid = en_grid,
  metrics = metric_set(roc_auc, accuracy)
)

autoplot(en_tune)
```

## Random Forest

```{r rf-setup}
rf_mod <- rand_forest(mtry = tune(), trees = tune(), min_n = tune()) %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("classification")

rf_wflow <- workflow() %>%
  add_model(rf_mod) %>%
  add_recipe(ipo_recipe)

# The recipe produces 29 predictor columns after dummy-coding (12 sector
# dummies + 2 usa_company + 6 employees_grouped + 4 ceo_gender + 5 numeric;
# confirmed with ncol(ipo_baked) - 1 above) so mtry can safely range up to the high teens.
rf_grid <- grid_regular(
  mtry(range = c(1, 15)),
  trees(range = c(200, 1000)),
  min_n(range = c(2, 20)),
  levels = 5
)
```

```{r rf-tune, eval=!models_cached}
rf_tune <- tune_grid(
  rf_wflow,
  resamples = ipo_folds,
  grid = rf_grid,
  metrics = metric_set(roc_auc, accuracy)
)

autoplot(rf_tune)
```

```{r rf-tune shutdown}
stopCluster(cl)
registerDoSEQ()
```

## Boosted Tree

```{r bt-setup}
bt_mod <- boost_tree(trees = tune(), tree_depth = tune(), learn_rate = tune()) %>%
  set_engine("xgboost") %>%
  set_mode("classification")

bt_wflow <- workflow() %>%
  add_model(bt_mod) %>%
  add_recipe(ipo_recipe)

bt_grid <- grid_regular(
  trees(range = c(100, 1000)),
  tree_depth(range = c(1, 10)),
  learn_rate(range = c(-3, -0.5)),
  levels = 4
)
```

```{r bt-tune, eval=!models_cached}
bt_tune <- tune_grid(
  bt_wflow,
  resamples = ipo_folds,
  grid = bt_grid,
  metrics = metric_set(roc_auc, accuracy)
)

autoplot(bt_tune)
```

```{r save-models, eval=!models_cached}
save(log_res, lda_res, en_tune, rf_tune, bt_tune,
     file = "data/ipo_tuned_models.rda")
```

# Model Selection

```{r compare-models}
best_log <- collect_metrics(log_res) %>% filter(.metric == "roc_auc") %>%
  mutate(model = "Logistic Regression")
best_lda <- collect_metrics(lda_res) %>% filter(.metric == "roc_auc") %>%
  mutate(model = "LDA")
best_en  <- show_best(en_tune, metric = "roc_auc", n = 1) %>% mutate(model = "Elastic Net")
best_rf  <- show_best(rf_tune, metric = "roc_auc", n = 1) %>% mutate(model = "Random Forest")
best_bt  <- show_best(bt_tune, metric = "roc_auc", n = 1) %>% mutate(model = "Boosted Tree")

model_comparison <- bind_rows(best_log, best_lda, best_en, best_rf, best_bt) %>%
  select(model, mean, std_err) %>%
  arrange(desc(mean))

model_comparison
```

```{r compare-plot, fig.width=6, fig.height=4}
ggplot(model_comparison, aes(x = fct_reorder(model, mean), y = mean)) +
  geom_col(fill = "#2c7fb8") +
  geom_errorbar(aes(ymin = mean - std_err, ymax = mean + std_err), width = 0.2) +
  coord_flip() +
  labs(title = "Cross-Validated ROC AUC by Model Type",
       x = NULL, y = "Mean ROC AUC (10-fold CV)") +
  theme_minimal()
```


```{r finalize}
best_params <- select_best(rf_tune, metric = "roc_auc")
final_wflow <- finalize_workflow(rf_wflow, best_params)
final_fit   <- fit(final_wflow, data = ipo_train)
```

## Variable importance

```{r vip, fig.width=6, fig.height=4}
vi_scores <- final_fit %>% extract_fit_parsnip() %>% vip::vi()
vi_scores

ggplot(vi_scores, aes(x = fct_reorder(Variable, Importance), y = Importance)) +
  geom_col(fill = "#2c7fb8", color = "black") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Variable Importance, Final Model", x = NULL)
```

# Performance on the Testing Set

```{r test-performance}
ipo_test_res <- augment(final_fit, new_data = ipo_test)

test_roc_auc <- roc_auc(ipo_test_res, truth = pop, .pred_Pop)
test_acc     <- accuracy(ipo_test_res, truth = pop, estimate = .pred_class)

bind_rows(test_roc_auc, test_acc)
```

```{r test-roc-curve, fig.width=5, fig.height=5}
roc_curve(ipo_test_res, truth = pop, .pred_Pop) %>%
  autoplot() +
  labs(title = "ROC Curve, Final Model (Testing Data)")
```

```{r test-confusion, fig.width=5, fig.height=4}
conf_mat(ipo_test_res, truth = pop, estimate = .pred_class) %>%
  autoplot(type = "heatmap") +
  labs(title = "Confusion Matrix, Final Model (Testing Data)")
```

I compare `test_roc_auc` above to the mean cross-validated ROC AUC for the
same model in `model_comparison`. If the two numbers are close together
that's good evidence the cross-validation actually gave an honest estimate
of how the model generalizes. If the testing ROC AUC is noticeably lower,
that suggests some overfitting to the training folds. If it's noticeably
higher, that's most likely just because the testing set is relatively
small, not because the model is actually better than what CV suggested.

# Conclusion

From the five models I fit, the cross-validated ROC AUC clustered pretty tightly
in the 0.69-0.71 range no matter which model type I used. Random Forest ended up 
barely on top with a cross-validated ROC AUC of 0.710, just barely beating the 
boosted tree (0.703). The elastic net, LDA, and logistic regression were close behind 
at 0.693, 0.691, and 0.689. On the testing set, the final random forest scored a ROC AUC 
of 0.696 and an accuracy of 0.765 which is close to its 0.710 CV estimate. So 
cross-validation gave a pretty honest read on how the model would actually generalize. 
I think the fact that a simple logistic regression does almost as well as a tuned random forest 
is actually pretty interesting. It shows that the signal in this data is fairly linear
and additive instead of being driven by complicated interactions between predictors.
So the more flexible models don't really have much of an advantage here.

The variable importance plot is actually pretty important here. `year_difference`
and the four market-trend variables took the top five importance spots whcih were
all clustered pretty close together roughly 154 to 168. `sector_unknown` 
which was basically whether sector was actually missing for a given IPO came in
6th place at 112.8 This was a real drop from the top five, but still way ahead of 
every actual named sector like employee-count bucket or CEO-gender category.
This shows that market timing and company age matter the most but simply not having
a sector on record for a company is still a surprisingly strong signal on its own. 
Lookibg back at the sector EDA plot, IPOs with an unrecorded sector only popped 
about 1% of the time compared to roughly 15% overall. So the "we don't have this 
company's basic info" is a genuinely strong signal by itself.This is probably because 
the same smaller abd lower-profile listings that never got fully documented are 
also the ones least likely to get investors as interested on release day which makes sense.

With this the model is definitely far from perfect which I don't think that's really surprising.
Whether an IPO pops on its first day is driven by a lot of other uncalculated variables like 
investor sentiment, news about the similar stocks in the sector, and how the 
underwriters behave in the days leading up to the IPO. These numbers were not calculated
in this dataset and can drive the IPO's value. I think a model with a moderate 
but real ROC AUC is basically the expected outcome for a problem noisy modeling approach.

The class imbalance ended up mattering a lot here. My final model's confusion matrix 
on the testing set caught 40 of the 110 real pops and missed 70.It also had 97 false positives
giving an overall accuracy of 76.5%. That recall wouldn't have happened without step_upsample()
in the recipe with only about 15% of training rows labeled "Pop," Upsampling forces 
the model to actually learn the minority class during training, which is exactly why 
I leaned on ROC AUC and not accuracy to pick the model in the first place.

In the future I think the best steps would be adding more features instead of trying even more model types.
The raw file actually has hundreds of pre-IPO financial-statement line items like 
revenue, net income, total assets and more that I left out here because they're extremely
specific to certain industries. Building a version of this model around just a handful of these
especially total revenue and net income the year before IPO would be the next thing to try. 
I also think fitting the model to more specfic sectors which would make these predictors
more useful. Also training the model for more specfic IPO's like health would make these models
more accurate. 

In conclusion, calculating if an IPO would "pop" or not is a very hard thing to model since
it is so noisy. If it was this easy to calculate every big prop firm would do it. 
I think this was a very interesting project to do really showing how much goes into predicting stock
direction which is why big firms pay so much and why quants make so much. In the end
it was a fun model to train even though the recall rate was only 36.4% and 29.2% precision
with a ROC of ~0.7, which wouldn't be actually used to find if an IPO would "pop" or not.






